Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-08-03 manuscript.rmd
symbol edits

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

- Deep water access is believed to be an important, drought-tolerance mechanism, but under-studied given the lack of data. Vulnerability to embolism ($\Psi_{88,~stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test the role of deep-water access in drought-tolerance strategies of a hyper-diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$) that scaled new data on 21 species with Leaf Mass Area and Wood Specific Gravity. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, but lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for predicting the future of tropical forests under climate change.

@@

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak:2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

@@

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions). A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak:2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting or water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

@@

\
Here, for large trees of length(erd.sp) canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 34 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) whose soil water dynamics ($\Psi_{~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

@@

\
In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit:2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.

@@

We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ and $LMA$, defined as mean wood specific gravity after drying at 100$^\circ$C (gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (gm^-2^) for leaves receiving direct sunlight, respectively.  Fitted polynomial equations are described as,

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$ (cm.yr^-1^)) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx). 
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs. \@ref(eq:M2)- \@ref(eq:M4).

@@

For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (%~$\circ$~; *n* = 7). We used Xylem sap $\delta^2H$ values given in [@Meinzer:2001] for the dry-season (March 1997) as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil. We removed species that are deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced in March and thus may not comparable to other species with leaves present. Fig. S\@ref(fig:iso-data). We also removed *Guapira* *spp*, whose xylem sap isotope value was not constrained by the soil water isotope depth profile.

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Fig. (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yrperiod for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

# Discussion
\
Our analyses revealed that co-occurring large canopy tree species (*n* = ` r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk. At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015; @Condit.2017)

@@

Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, that are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015]. Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak:2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought. Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak:2018 and this study]. We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality. For example. based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].

and the usefulness of looking at species ecological strategies for that. As those are developed in response to plants historical environment. Plant traits hold clue for the historical environment that shaped their ecological strategy [@Reich:2014].

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak:2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$. Model by @Chitra-Tarak:2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves. For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI. They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters). Our inverse ERD model fits only two parameters by way of regressing depth-specific model against 5 growth data points. Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references]. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?]. We also note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation. Model by Chitra-Tarak et al @Chitra-Tarak:2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers. comm.). Our model of belowground water-availability (ELM-FATES) was also advanced compared to that used by Chitra-Tarak et al [2018] and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge, unlike theirs, which was calibrated on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.

>>>>

---
title: "Drought-sensitive trees survive on deep water access in a tropical forest community"
author: ''
date: ''
output:
  bookdown::html_document2:
    number_sections: no
    theme: flatly
    fig_caption: yes
  bookdown::word_document2:
        reference_docx: manuscriptstyle.docx
        fig_caption: yes
  bookdown::pdf_document2: default
bibliography: references.bib
csl: new-phytologist.csl
---

@@

- Deep water access is believed to be an important, drought-tolerance mechanism, but under-studied given the lack of data. Vulnerability to embolism ($\Psi_{88,~stem}$) and safety margins ($\Psi_{min} - \Psi_{88,~stem}$) are shown to be predictors of drought-induced mortality risk, but deep water access may delay plants from reaching these critical thresholds. Here we test the role of deep-water access in drought-tolerance strategies of a hyper-diverse tropical forest community (Barro Colorado Island, Panama) using a novel data-model approach.
- We inversely modelled Effective Rooting Depth (ERD) for large trees of `r length(erd.sp)` canopy tree species by linking species growth dynamics (6 censuses, 1990-2015) to VPD, Soil Water Potential ($\Psi_{soil}$) by depth in the whole-soil column, and a new community-scale model of species vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$) that scaled new data on 21 species with Leaf Mass Area~lamina~ and Wood Specific Gravity~100^$\circ$^C~. We evaluated ERD by leveraging existing isotopic tracer data ($\delta^2H$) on tree xylem sap and soil column. 
- We find that deeper ERD is associated with higher stem hydraulic conductivity, greater vulnerability to xylem embolism, narrower safety margins, but lower mortality rates over 35 years. Deeper ERD species spent less time below critical hydraulic thresholds indicating that ERD buffers drought-stress.
- The role of deep water access in alleviating mortality of drought-sensitive trees have important implications for predicting the future of tropical forests under climate change.

@@

# Introduction
\
Intensifying droughts threaten tropical forests, a biome that plays a disproportionately large role in the global carbon and energy cycle, and supports half of the global biodiversity [@Bonan:2008; @Wright:20050th]. Tropical forests are considered an especially drought-vulnerable biome given the combination of climate risk and vegetation sensitivity [@Meir:2015trw]: here, historic climatic variation is relatively small [@Mora.2013], and the observed ranges of trees' hydraulic safety margins are relatively narrow [@Choat:2012; @Oliveira:2019]. Widespread drought-induced mortality has already been observed across the tropics for specific tree groups [@Bennett:2015; @Chitra-Tarak.2018; @Hilker:2014; @Phillips:2010]. However, state-of-the-art dynamic vegetation models (DVMs) fail to capture drought-induced vegetation dynamics in the tropics [@Galbraith.2010; @Powell:2013], as underlying mechanisms of drought-tolerance are not fully understood.

@@

\
Establishing relationships between plant's environment, traits and demographic outcomes is imperative for a predictive plant ecology [@Mcgill.2006], but plant's realised water environments are rarely studied. Huge efforts are devoted to linking the same-for-all-trees drought indices, above-ground traits and demographic rates. These efforts have resulted in identification of above-ground traits such as proxies of hydraulic failure as indicators of mortality risk. A recent meta-analysis found greater drought-induced mortality rates in species with cavitation-vulnerable xylem ($\Psi_{88,~stem}$) and narrow safety margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Anderegg:2016], but counter examples also exist [@Hoffmann:2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013]; indicating whether plants reach critical thresholds of hydraulic failure depends on their realised drought intensity (see Table \@ref(tab:symbols) for trait symbols and their definitions). A rare study that accounted for realised water availability for four tree species in a centennial drought (Texas, USA) found that mortality decreased with deep rootedness even though deep rooted species were more vulnerable to xylem cavitation (high $\Psi_{88,~stem}$) and had narrower safety-margins ($\Psi_{min} - \Psi_{88,~stem}$) [@Johnson:2018fad]. A species sourcing deep water in a Karst system (Italy) showed lower hydraulic failure and mortality in an extreme drought despite having high $\Psi_{88,~stem}$ and narrower safety margin compared to two species sourcing shallow-water [@Nardini:2015]. @Chitra-Tarak.2018 proposed the first community-scale model that linked trees' realised water availability to demographic outcomes through droughts. They found that species accessing deep-water in a tropical dry forest (India) died more in a rare, centennial drought compared to shallow-water species. They concluded that this was consistent with an 'aquisitive-conservative' species strategy spectrum. Aquisitives relied on stable water resource at depth, which exhausted in the multi-annual drought, resulting in reduced survival, whereas conservative species used fluctuating shallow-water and survived the drought as some shallow-water was available via infiltration. However, lack of data on rooting or water-sourcing depths and hydraulic traits precluded this study from corroborating modelled estimates of water-uptake depths, or investigating trait coordination.

@@

\
Here, for large trees of length(erd.sp) canopy species of a hyperdiverse neo-tropical moist forest (BCI, Panama) we make a thorough assessment of trade-offs between belowground water environment, above-ground hydraulic traits, estimates of effective rooting depths and demographic outcomes over 34 years that withnessed droughts of various intensity and duration. We obtained daily dynamics of plant water potential ($\Psi_{soil,~z}$) by depth $z$ in the whole-soil column with a locally parametarised and calibrated 1-D hydrological water balance model. Developing a novel inverse model allowed us to estimate species effective rooting depths (ERD), defined as the depth ($z$) at which soil water dynamics ($\Psi_{soil,~z}$) best explained observed dynamics of species growth (six 5-yearly censuses, 1990-2015). This model incorporated impact of atmospheric drought (VPD) on growth and was constrained by above-ground hydraulic traits, namely, species leaf vulnerability curves ($K_{leaf}$ vs. $\Psi_{leaf}$). We collected new data on vulnerability curves for xx species. Identifying scaling relationships between these curves and commonly available traits (SLA, specific wood density) allowed us to scale-up these curves to `r length(data.model.AB$sp)` tree species in the community. We evaluated ERD against existing data on plant and soil water isotopes ($\delta^2H$) and coordination of ERD with hydraulic traits and mortality rates.

@@

\
In the 50ha plot all stems above 10 mm diameter at breast height (dbh) are mapped, tagged with a unique number, identified to species level and measured every 5 years through 1985-2015 for growth, mortality as well as recruitment of new stems into the 10 mm size class [@Condit.2017]. This inventory represents 321 woody species across the eight censuses. Aside from a 2ha patch of ~100 year old secondary forest at the north-eastern corner of the plot the remaining 48 ha represent old growth forest that was never clear cut for agriculture. The plot (100 m x 50 m) was originally chosen for its relative uniformity of relief and other physical conditions; the elevation range is only 40 m (120-160 m) [@Harms:2001]. The plot covers most of the upper catchment of Conrad Creek. The shallow valley of this creek is the main interruption to the otherwise mostly flat or gently sloping plot. The valley gives rise to the 1.5-ha swamp at the end of the wet season. Soils of the plot are homogenous with red light clays, described as the Ava soil type, accounting for 72.4% of the plot (BCI soil survey report). The topsoil field texture is silty clay loam that gradually fines to silty clay in the subsoil. Soil is mostly free draining, but restricted subsoil permeability gives rise to temporary wet season ponding. Detailed descriptions of the climate, geology, flora and fauna of BCI can be found elsewhere.

@@

We identified scaling relationships between fitted parameters $A$ and $B$ in Eq. \@ref(eq:vc-exp) and two traits that were available at community-scale, namely, $WSG$ (or, WSG~100^$\circ$^C~) and $LMA$ (or, LMA~lamina~), defined, respectively, as mean wood specific gravity after drying at 100$^\circ$C (gcm^-3^) and mean leaf mass per unit area measured for the leaf lamina excluding the petiole and for compound leaves the petiolules (gm^-2^) for leaves receiving direct sunlight.  Fitted polynomial equations are described as,

@@

Growth Data: To minimise effect of light variation among trees and census intervals we selected only large trees (>= 30 cm dbh, diameter at breast height) and species whose maximum height was at least 30 m (hereafter, canopy species). We chose only those trees for which diameter (dbh) record was present across all size censuses (1990, 1995, 2000, 2005, 2010, 2015). We calculated individual tree ($i$) growth rates ($Gr_{i,t}$; cm.yr^-1^) for each of the five census intervals ($t$) as the difference between tree dbh measured in successive censuses, divided by the inter-census interval based on the exact date of measurements. As tree growth rates increase with size, to keep growth rates comparable across census intervals, we obtained residuals from a dbh model of tree growth based on a tree's dbh at the beginning of a census interval ($Resid_{i,t} = Gr_{i,t} - f(DBH_{i, t})$). The diameter model of growth was fitted based on all trees' data pooled together as there was insufficient species-specific data for large trees (SI ). We then standardized (centered and scaled) an individual tree time series ($Resid_{i,t}$) and obtained species-level (*n* = `r length(erd.sp)`) times-series of medians ($Growth_{t}$) that had data on at least three trees (median = xx). 
As VPD may have a non-linear relationship with growth, we fitted a polynomial curve between GPP (gCm^2^day^-1^) and VPD (kPa; $GPP = f(VPD) + error$;  SI) and used standardised $f(VPD)$ (range 0-1) as $StdVPD$ in Eqs. \@ref(eq:M2)- \@ref(eq:M4).

@@

For each model, species ERD was regressed against species Xylem sap $\delta^2H$ (%~$\circ$~; *n* = 7). We used Xylem sap $\delta^2H$ values given in [@Meinzer:2001] for the dry-season (March 1997) as these showed largest seasonal divergence among species, with a corresponding gradient of $\delta^2H$ in the soil. We removed species that are deciduous in March as xylem sap $\delta^2H$ of these species may not be linked with water sourced in March and thus may not comparable to other species with leaves present. Fig. S\@ref(fig:iso-data). We also removed *Guapira* *standleyana*, whose xylem sap isotope value was not constrained by the soil water isotope depth profile.

@@

\
Species with deeper $ERD$ showed lower mean mortality rates across the 34-yr record (1982-2015; *R*^2^ = 0.11, *P* = 0.075, *n* = `r length(erd.mrate.sp)`; Fig. \@ref(fig:erd-mort)). For five out of seven census intervals, species with shallower $ERD$ had significantly higher mortality anomalies than deeper ERD species; Fig. (\@ref(fig:erd-mort-anom)). Analysis of time spent under catastrophic soil drought showed that the proportion of days species $\Psi_{ERD}$ was more negative than species $\Psi_{crit}$ decreased with species ERD (Fig. \@ref(fig:erd-time)). Species with ERD deeper than 1 m did not cross species $\Psi_{crit}$ for the entire 25-yr period for which soil water potential $\Psi_{soil,~z}$ was modelled (1990-2015), as soil depths below 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dry seasons and drought periods, which is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa).

@@

# Discussion
\
Our analyses revealed that co-occurring large canopy tree species (*n* = `r length(erd.sp)`) with deeper effective rooting depths (ERD) were associated with higher hydraulic efficiency, but not safety: higher maximum stem hydraulic conductance, but greater xylem vulnerability to embolism and narrower hydraulic safety margins compared to shallow ERD species. Yet, deeper ERD species showed lower mortality rates than shallower ERD species over the 34-yr record (1982-2015). Analysis of time spent below ($\Psi_{crit}$) in the soil zone matching ERD $\Psi_{soil,~z = ERD}$ showed that this was likely due to decreasing realised risk of hydraulic failure with depth. We thus show, for the first time, that interspecific variation in trees' effective rooting depths aligns species on a 'fast-slow' water economic spectrum by demonstrating a link between trees' realised drought experience, hydraulic architecture and long-term mortality outcomes. On the 'fast' end of the spectrum were deep ERD species with hydraulically efficient, but risky suit of traits optimised for an environment of reliable and abundant source of water at depth that actualised to a lower mortality-risk. At the slow end of the spectrum were shallow ERD species with inefficient, but safer xylem, likely optimised for high variability in $\Psi_{soil}$ in the shallow soil layers, resulting in higher variance in mortality rates across censuses as $\Psi_{soil,~z = ERD}$ crossed ERD $\Psi_{crit}$. Mortality rates of shallow ERD species may have been exacerbated by increased drought events during the study period. Previous work shows that half of the extreme dry seasons, estimated as < 1 SD annual water deficit, during 1931-2015 (85-yr) occurred during 1982-2005 (24-yr) of the 34-yr long study period (1982-2015; @Condit.2017)

@@

Our study brings attention to the need for assessing drought-sensitivity in terms of species realised drought risk. We found that species accessing deeper water had greater xylem vulnerability to embolism and narrower branch hydraulic safety margins traits, that are identified as proxies for mortality risk, but they were less droughted and had lower mortality compared to species relying on shallower water through a study period of greater recurrence of drought. These results are consistent with the rare studies that studied three or four species each with known rooting or water sourcing depths and assesed their hydraulic traits and mortality rates during a centennial (Texas) or extreme (Italy) drought [@Johnson:2018fad; @Nardini:2015]. Our result of lower mortality in deep rooted trees contrasts with the finding of @Chitra-Tarak.2018, who found that species identified by their inverse ERD model as effectively deep rooted died *more* in a prolonged, centennial drought. Their hydrological model found that in the multi-annual drought deep water was exhausted for over two years, whereas our hydrological modeling at BCI finds that despite higher frequency of droughts during our study period $\Psi_{soil}$ below the depth of 1 m remained above `r round(psi.1.7.min$median, 2)` (`r round(as.numeric(psi.1.7.min$q97.5), 2)`, `r round(as.numeric(psi.1.7.min$q2.5), 2)`) (median, 95% CI) through all the dryseasons and drought periods, that is above ($\Psi_{crit}$) of the most sensitive tree species (-0.28 MPa). The difference in our results are likely attributable to site differences in the precipitation regime. In their dry forest site (Mudumalai National Park, India) precipitation is low (1095 $\pm$ 296 mm; mean $\pm$ SD) compared to BCI (2687 $\pm$ 515 mm), and precipitation to potential evapotranspiration ratio at Mudumalai is nearly one (PET = 980 $\pm$ 64 mm), whereas at BCI it is nearly two (PET = 9xx $\pm$ xx mm), and thus at BCI deep soil layers are effectively recharged annually [@Chitra-Tarak.2018 and this study]. We note that the variation in mortality explained by in our study is low (*R*^2^ = 0.11, *P* = 0.075), and other causes may explain observed mortality. For example, based on a year-long study, lightening explained ~4.5% of mortality for trees >=10 cm dbh and ~40% for trees >=60 cm dbh [@Yanoviak:2019].

The trade-offs that we identify between above- and below-ground traits, water environment vertically belowground and mortality are important for representing diversity in Dynamic Global Vegetation Models especially in species-rich forests. The hydraulic efficiency-safety trade-off between high values of maximum leaf or stem conductivity (high $K_{max}$) and cavitation-resistance (low $\Psi_{50}$ or $\Psi_{88}$) is hypothesised although community-wide examples are rare [@Hacke:2006; @Gleason:2016]. We report such a trade-off at both leaf (*n* = 21 species) and branch level (*n* = 27 species, but data measured on trees in SLZ and PNM, only include those species that are found in BCI?; Or shall I add those species numbers and a SI figure for only those species for which these data are available for both leaf and branch?). We also found that LMA~lamina~ and WSG~100^$\circ$^C~ were strong predictors of PLC curves for leaf, but not branch (add SI figures). High $K_{max}$ is elsewhere associated with photosynthetic rates with faster growth potential, but we did not find a relationship between $K_{max}$ and growth rates (*n* = 21 species).

### Effective rooting depths at the tree community level
\
By estimating ERD for large trees of `r length(erd.sp)` canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. Our chosen ERD model is an advancement over the only other inverse model of ERD (or water uptake depth index, @Chitra-Tarak.2018, *n* = 12 species) in terms of physiologically meaningful representation, fewer parameters, corroboration with tree and soil data on $\delta^2H$. Model by @Chitra-Tarak.2018 estimated six species-specific phenomenological parameters: three as root fraction for corresponding lumped vadose zone layers, and three as shape parameters for water uptake vs availability curves. For each species, they calculated daily water uptake successively from three layers--the supply, to satisfy the demand of the daily maximum evapo-transpiration, a function of PET and LAI. They regressed total supply to total demand ratio over census intervals (termed as growth coefficient) against growth (5 data points), and chose values for six parameters that best-fit (highest R^2^, that is, 2 additional parameters). Our inverse ERD model fits only two parameters by way of regressing depth-specific model against 5 growth data points. Our model was constrained on VPD and species leaf vulnerability (PLC) curves, variables that are shown to be physiologically meaningful [add references]. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (LMA~lamina~ and WSG~100^$\circ$^C~) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?]. We also note the availability of data on $\delta^2H$ on trees and water sources belowground--soil/substrate matrix water in the vadoze zone or free water below the water table--as important for model calibration/validation. Model by @Chitra-Tarak.2018 were not corroborated with such data as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Jean Riotte, pers. comm.). Our model of belowground water-availability (ELM-FATES) was also advanced compared to that used by @Chitra-Tarak.2018 and was calibrated with data on soil moisture dynamics by depth (15, 40, 100 cm), as well as, evapotranspiration and stream discharge, unlike theirs, which was calibrated on stream discharge and water table levels. We lacked data on water table levels which would have been useful to constrain water availability dynamics in the unobserved soil layers.
